### 2016-06-27 publications/Journal/3-features.tex
Update section 3.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

We will not make a detailed comparison with PythonPDEVS here, but only mention some supported features.
Dxex supports, similarly to PythonPDEVS, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols (though only in parallel).
This is in line with the design principle of PythonPDEVS: allow users to pass performance hints to the simulation kernel.
In our case, a user can pass the simulation kernel the synchronization protocol to use for this model.
Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language. Dxex will use new optimizations from the C++14 standard when possible.

@@

\subsubsection{Conservative}
For conservative synchronization, each node must determine the nodes it is influenced by.
Each model needs to provide a lookahead function, which determines the lookahead depending on the current simulation state.
Within the returned time interval, the model promises not to raise an event.
A node aggregates this information to computes its earliest output time (EOT).
This value is written out in shared memory, where it can be read out by all other nodes.

@@

@@

Like PythonPDEVS dxex has full tracing capabilities supporting a wide number of output formats : Json, XML, Cell, console, ... . The user can implement his own outputpolicy and use it in our modular tracing design. The tracing functionality itself is implemented asynchronous and thread-safe. By overriding the output functions in the Message or State classes the user can generate model specific output for use with an existing tracing outputpolicy. Tracing can be completely disabled to maximize simulation performance should this be desired (e.g. benchmarks). 
% Talk about adevs?

@@

As we will demonstrate in ~\nameref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulation with high event frequency, especially in parallel. Dxex has a modular allocation interface that separates the object creation in kernels and models from the actual implementation. This allows us by using a single compilation parameter to change the allocation strategy per kernel type. This interface allows instrumentation of memory usage. Not only useful to detect memory leaks should this be a concern, it can also be of use to a model author to trace event creation in large models. A model author can rely on the default implementation as the fastest common configuration verified by the benchmarks.

@@

A model author can specify which kernel a model should be allocated to, should such manual intervention be required. This is handled by the default model allocator. If no preference is given a simple striping scheme is followed. It should be obvious that this is not sufficient in most simulations. In Section \nameref{sec:4-performance} we analyze allocation schemes and their performance impact. By overriding the default allocator a model author can tune the allocation scheme to maximize parallel speedup. This interface can be in future work also be used to employ graph algorithms for an automatic allocation scheme, for example one that avoids cycles in the resulting kernel dependency graph.

@@

Tracing offers a black box view of a DEVS model in a simulation, with transitions, events and states recorded at each state change. Dxex supplements this optionally with more in depth insights into the actual simulation. The allocation of models across kernels is visualized at the end of a simulation, with graphical representation of message frequency. This allows a user to fine tune future allocation schemes and the interface used by the visualization tool can also be used for more advanced purposes such as dynamic tracing of the simulation.\\
Invaluable during the development of dxex were runtime statistics collected by a simulation kernel. These record any measurable event in the simulation such as : events created/destroyed, inter kernel event frequency, reverts, GVT calculations, lookahead and eot values and insights into the fairness between the kernels themselves. Using this information it is possible to detect an imbalance between the simulation rounds that a single kernel executes compared to dependent or influencing kernels, or observe how balanced inter kernel message traffic actually is during a simulation. The same information can, as in the visualization tool, be used programmatically to extend dxex with visualization software that shows the simulation from a white box perspective or allow activity tracking.
Figures \ref{fig:pholdtree_visualize_seq},\ref{fig:pholdtree_visualize_parDFS},\ref{fig:pholdtree_visualize_parBFS} give an example of the modular visualization output generated by dxex after simulation the PholdTree model. These figures demonstrate the effect of allocation visually, the high inter kernels event count will degrade performance, as we will demonstrate in 
\nameref{PholdTreeallocation} . Dotted lines represent connections between models that did not see events exchanged, directed edges represents events with frequency as weight and color to distinguish inter versus intra kernel events.
% Add figure
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth, height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000.eps}
	\caption{Visualization of PholdTree (d=1,n=3,t=5000) sequential simulation.}
	\label{fig:pholdtree_visualize_seq}
\end{figure}
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4BFS.eps}
	\caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with breadth first allocation and 4 kernels.}
	\label{fig:pholdtree_visualize_parBFS}
\end{figure}
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4DFS.eps}
	\caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with depth first allocation and 4 kernels.}
	\label{fig:pholdtree_visualize_parDFS}
\end{figure}

>>>>

We will not make a detailed comparison with PythonPDEVS here, but only mention some supported features.
Dxex supports, similarly to PythonPDEVS, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols (though only in parallel).
This is in line with the design principle of PythonPDEVS: allow users to pass performance hints to the simulation kernel.
In our case, a user can pass the simulation kernel the synchronization protocol to use for this model, or even switch the synchronization protocol during runtime.
Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language. Dxex will use new optimizations from the C++14 standard when possible.

@@

\subsubsection{Conservative}
For conservative synchronization, each node must determine the nodes it is influenced by.
Each model needs to provide a lookahead function, which determines the lookahead depending on the current simulation state.
Within the returned time interval, the model promises not to raise an event.
A node aggregates this information to compute its earliest output time (EOT).
This value is written out in shared memory, where it can be read out by all other nodes.

@@

\subsubsection{Runtime synchronization switching}
A user can request a runtime switch between synchronization protocols. A typical use case is when a significant slowdown is observed or memory usage reaches a certain treshold. The simulation engine will swap out the current kernels with new instances while preserving the state of the models and the simulation itself, and resume using the new synchronization protocol. There is no real limit on the number of switches, but the replacing and synchronizing of the kernels comes at a runtime cost that is dependent on the number of present kernels and the actual hardware the simulation runs on. A test case requires for 2 kernels an overhead approaching a second. This is an experimental but promising feature that can be used in an automated framework driven by machine learning techniques.

@@

Like PythonPDEVS dxex has full tracing capabilities supporting a wide number of output formats : Json, XML, Cell, console. The user can implement a new outputpolicy and use it in our modular tracing design. The tracing functionality itself is implemented asynchronous and thread-safe, the user does not need to use synchronization primitives. By overriding the output functions in the Message and State classes the user can generate model specific output for use with an existing tracing outputpolicy. Tracing can be completely disabled to maximize simulation performance should this be desired (e.g. benchmarks). 

@@

As we will demonstrate in ~\nameref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulation with high event frequency, especially in parallel with optimistic synchronization. Dxex has a modular allocation interface that separates the object creation in kernels and models from the actual implementation. This allows us by using a single compilation parameter to change the allocation strategy. This interface allows instrumentation of memory usage. Not only useful to detect memory leaks should this be a concern, it can also be of use to a model author to trace event creation in large models. The default implementation is the fastest common configuration verified by the benchmarks used in this publication.

@@

A model author can specify which kernel a model should be allocated to, should such manual intervention be required. This is handled by the default model allocator. If no preference is given a simple striping scheme is followed but this is not sufficient in most simulations to achieve a speedup in parallel. In Section \nameref{sec:4-performance} we analyse the performance impact of allocation schemes. By overriding the default allocator a model author can tune the allocation scheme for a specific model to maximize parallel speedup. This interface can be in future work also be used to employ graph algorithms for an automatic allocation scheme, for example one that avoids cycles in the resulting kernel dependency graph.

@@

Tracing offers a black box view of a DEVS model in a simulation, with transitions, events and states recorded at each state change. Dxex supplements this optionally with more in depth insights into the actual simulation. 
The allocation of models across kernels is visualized at the end of a simulation along with the event trace of all events exchanged, including reverted events in optimistic. 
A user can then fine tune future allocation schemes.

Invaluable during the development of dxex were runtime statistics collected by a simulation kernel. These record any measurable event in the simulation such as : events created/destroyed, inter kernel event frequency, reverts, GVT calculations, lookahead and eot values and insights into the fairness between the kernels themselves. Using this information it is possible to detect an imbalance between the simulation rounds that a single kernel executes compared to dependent or influencing kernels, observe how balanced inter kernel message traffic actually is during a simulation or measure the reverted simulation transition in optimistic synchronization.
The same information can, as in the visualization tool, be used programmatically to extend dxex with visualization software that shows the simulation from a white box perspective or allow activity tracking.
Figures \ref{fig:pholdtree_visualize_seq},\ref{fig:pholdtree_visualize_parDFS},\ref{fig:pholdtree_visualize_parBFS} give an example of the modular visualization output generated by dxex after simulation the PholdTree model. These figures demonstrate the effect of allocation visually, the high inter kernels event count will degrade performance, as we will demonstrate in 
\nameref{PholdTreeallocation} . Dotted lines represent connections between models that did not see events exchanged, directed edges represents events with frequency as weight and color to distinguish inter versus intra kernel events.
% Add figure
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth, height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000.eps}
	\caption{Visualization of PholdTree (d=1,n=3,t=5000) sequential simulation.}
	\label{fig:pholdtree_visualize_seq}
\end{figure}
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4BFS.eps}
	\caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with breadth first allocation and 4 kernels.}
	\label{fig:pholdtree_visualize_parBFS}
\end{figure}
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth,  height=6cm, keepaspectratio]{fig/pholdtreed1n3t5000c4DFS.eps}
	\caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with depth first allocation and 4 kernels.}
	\label{fig:pholdtree_visualize_parDFS}
\end{figure}
